Measurement Bias from Address Aliasing

### 2016-01-09 paper.tex
Fix various smaller issues

Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
The issues they find are either handled explicitly or are irrelevant to our setup.
%They find intra-machine variations of up to 1.07\%,

@@

\section{4K Address Aliasing}
\label{sec:aliasing}
Modern processors are \emph{superscalar}, and achieve parallelism by issuing multiple instructions simultaneously and out of order.
One of the issues that can limit throughput is dependencies between a load and previous stores.
To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory these operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess} as well. 
%More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
Loads are then issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
The prediction is later verified, replaying any instructions that were wrongly assumed to have no dependencies.
Similarly, if the load and store locations are the same, the value can be \emph{forwarded} from the store before it retires.

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096 bytes.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
The number of times this happens can be counted by the following event:
\begin{description}
  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \hfill \\ \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
\end{description}
This event is listed in the manual for microarchitectures going back to ``Nehalem'', including ``Ivy Bridge'' and ``Haswell'' CPUs~\cite{Volume3B}.
Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:
\begin{description}
  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \hfill \\ \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
\end{description}
As address aliasing depends on the memory addresses of loads and stores, any environmental factor that affects memory layout has the potential to induce aliasing conditions.
In the following sections, we show how performance penalties from address aliasing can be the root cause of measurement bias.

@@

We see a huge change from the median in the number of alias events. 
%If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
We found that there are near zero address aliasing events everywhere except exactly the points we observe bias. 
The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from around 272,000 to 136,000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the 8 execution ports have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change.
The memory disambiguation system automatically assumes dependency between a load and a previous store that are aliased, which can limit the potential for issuing many instructions simultaneously and out of order~\cite[Page 2-20]{OptimizationManual}.
In our case, lower occupation of execution ports, and less pressure on the reservation station, can be explained by the CPU not being able to issue more operations while waiting for an aliased memory access to be resolved.

@@

\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
\pgfplotstableread{bin/conv-default-o3-haswell.estimate.dat}{\convtablethreehaswell}
\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=Haswell -O2,
        font=\footnotesize,
        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.05, % Make it fit to text width side by side
        height=6.0cm,
        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \hspace{0.5cm}
  \begin{tikzpicture}
    \begin{axis}[
        title=Haswell -O3,
        font=\footnotesize,
        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.05,
        height=6.0cm,
        skip coords between index={20}{32}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethreehaswell ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethreehaswell ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \hspace{0.5cm}
  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}. Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
\end{figure*}

@@

%\newpage
\section{Aliasing in Real Applications}
\label{sec:blas}
To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
\iffalse
The functionality is divided into three categories:
\begin{description}
  \item[Level 1] \hspace{0.1cm} Scalar and vector operations, such as dot product and vector addition.
  \item[Level 2] \hspace{0.1cm} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
  \item[Level 3] \hspace{0.1cm} Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
\end{description}
Procedures operating on vectors, intuitively appears the most likely to have potential for aliasing.
\fi
Here, the level 1 and 2 vector operations intuitively appears to be more likely to have potential for aliasing.
We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
%In code, the function is invoked with the following parameters:
%\begin{lstlisting}[breaklines=true]
%    cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
%\end{lstlisting}

>>>>

%Reliability of PMU measurements have been evaluated by Weaver and Mckee~\cite{Weaver:2008:PCTrusted}.
%The issues they find are either handled explicitly or are irrelevant to our setup.
%They find intra-machine variations of up to 1.07\%,

@@

\section{4K Address Aliasing}
\label{sec:aliasing}
Modern processors are \emph{superscalar}, and achieve parallelism by issuing multiple instructions simultaneously and out of order.
One of the issues that can limit throughput is dependencies between a load and previous stores.
To increase parallelism, modern architectures use a technique called \emph{memory disambiguation} to execute memory operations out of order~\cite{Intel:2006:InsideICM:SmartMemoryAccess}.
%More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
Loads are issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store that is still not retired.
The prediction is later verified, replaying any instructions that were wrongly assumed to have no dependencies.
Similarly, if the load and store locations are the same, the value can be \emph{forwarded} from the store before it retires.

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096 bytes.
%A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
%Despite being independent, 
In these cases, the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
The number of times this happens can be counted by the following event:
\begin{description}
  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \hfill \\ \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
\end{description}
This event is listed in the manual for microarchitectures going back to ``Nehalem'', including ``Ivy Bridge'' and ``Haswell'' CPUs~\cite{Volume3B}.
Older architectures such as Core do not have this particular event listed, but 4K aliasing is covered by a more general event:
\begin{description}
  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \hfill \\ \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
\end{description}
As address aliasing depends on the memory addresses of loads and stores, any environmental factor that affects memory layout has the potential to induce aliasing conditions.
In the following sections, we show how performance penalties from address aliasing can be the root cause of measurement bias.

@@

We see a significant change from the median in the number of alias events. 
%If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
We found that there are near zero address aliasing events everywhere except exactly the points we observe bias. 
The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from around 272,000 to 136,000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the 8 execution ports have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change.
The memory disambiguation system automatically assumes dependency between a load and a previous store that are aliased, which can limit the potential for issuing many instructions simultaneously and out of order~\cite[Page 2-20]{OptimizationManual}.
In our case, lower occupation of execution ports, and less pressure on the reservation station, can be explained by the CPU not being able to issue more operations while waiting for an aliased memory access to be resolved.

@@

\pgfplotstableread{bin/conv-default-o2-haswell.estimate.dat}{\convtabletwohaswell}
\pgfplotstableread{bin/conv-default-o3-haswell.estimate.dat}{\convtablethreehaswell}
\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        title=Haswell -O2,
        font=\footnotesize,
        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.05, % Make it fit to text width side by side
        height=6.0cm,
        skip coords between index={20}{32} % Limit level of detail to fit page width nicely
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwohaswell ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwohaswell ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \hspace{0.5cm}
  \begin{tikzpicture}
    \begin{axis}[
        title=Haswell -O3,
        font=\footnotesize,
        xlabel=Relative offset in \texttt{sizeof(float)} bytes,
        ylabel=Event count,
        cycle list name=black white,
        width=\textwidth/2.05,
        height=6.0cm,
        skip coords between index={20}{32}
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethreehaswell ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethreehaswell ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \hspace{0.5cm}
  \caption{\label{fig:conv-default}Cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Figure~\ref{lst:conv}, input size $n=2^{20}$.}
  %Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Showing results for optimization levels O2 and O3 compiled with GCC, input size $n=2^{20}$.}
\end{figure*}

@@

%\newpage
\section{Aliasing in Real Applications}
\label{sec:blas}
To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
\iffalse
The functionality is divided into three categories:
\begin{description}
  \item[Level 1] \hspace{0.1cm} Scalar and vector operations, such as dot product and vector addition.
  \item[Level 2] \hspace{0.1cm} Matrix-vector operations, such as gemv for general matrix-vector multiplication.
  \item[Level 3] \hspace{0.1cm} Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
\end{description}
Procedures operating on vectors, intuitively appears the most likely to have potential for aliasing.
\fi
%The level 1 and 2 routines involving vectors intuitively appears to be more likely to have potential for aliasing.
We look at three different libraries available through the Ubuntu package manager; \emph{libblas}, \emph{libatlas}, and \emph{libopenblas}~\cite{Whaley:1998:ATLAS,Wang:2013:OpenBLAS}.
We ran our bias analysis on the level 2 procedure \texttt{cblas\_dgemv}, which computes the matrix-vector product $\boldsymbol{y} = \alpha\text{op}\left(A\right)\boldsymbol{x} + \beta\boldsymbol{y}$ for \texttt{double} sized numbers.
In this equation,  $\alpha$ and $\beta$ are constants, and $\text{op}\left(A\right)$ is an optional transpose or complex conjugate of the matrix.
We set $\alpha = 1$, $\beta = 0$ and $\text{op} \left(A\right) = A$ to reduce the formula to $\boldsymbol{y}=A\boldsymbol{x}$.
Denote the matrix size as $M \times N$, making $\boldsymbol{x}$ of length $N$ and $\boldsymbol{y}$ of length $M$.
%In code, the function is invoked with the following parameters:
%\begin{lstlisting}[breaklines=true]
%    cblas_dgemv(CblasColMajor, CblasNoTrans, M, N, alpha, A, M, x, 1, beta, y, 1);
%\end{lstlisting}
